Hybrid quantum-classical unsupervised data clustering based on the Self-Organizing Feature Map

### 2024-11-11 manuscript/main.tex
Remove trailing whitespace
Replace distance matrix figures with one that used 4 data qubits and extra averaging

%%%%%
\begin{figure}[t]
	\includegraphics[width=\columnwidth]{qcircuit.eps}
    \caption{
		Quantum circuit for the quantum parallelized Hamming distance calculation between all pairs of binary vectors from two sets ${X}$ and ${Y}$.
		First, we encode information about pairwise different qubits in a quantum state of the $X$-register by applying the CNOT gates.
		Second, the Hamming distance values are extracted to the amplitudes of superposition with the control phase rotation (see Eq.~(\ref{eq:controled_phase_rotation})) and Hadamard gates.
		Finally, a quantum space of the $X$-register is returned to the initial basis for information retrieval.
	}
	\label{fig:qcircuit}
\end{figure}
%%%%%

%%%%%
\begin{figure}[t]
	\includegraphics[width=0.95\columnwidth]{distance_matrix.png}
	\caption{
		Hamming distance matrix between two data sets of binary vectors.
		Values of distance that less than median distance value was marked black.
		The classical simulation of the quantum circuit shows perfect agreement to theoretical calculations and presented on the left figure.
		Result obtained on  IBMQ ``ibmq\_16\_melbourne'' backend is shown on the right figure and the accuracy in comparison with the classical simulation is more than 92 $\%$.
	}
	\label{fig:distance_matrix}
\end{figure}
%%%%%

@@

Measuring the Hamming distance of a particular pair of input vectors $\left| x_i \right\rangle$ and cluster vectors $\left| y_j \right\rangle$ consists of extracting the relevant amplitude from the subspace that those states form,
this can be done using the following projection operator
%
\begin{align}
\Pi_{i,j} = &\left| x_i \rangle\langle x_i \right| \otimes \left| y_j \rangle\langle y_j \right| \otimes I .
\end{align}
%
Using the above projection operator, the subspace of the Hilbert space formed by a particular pair of input and cluster vectors can be traced out as
%
\begin{align}
    \rho_{i,j} &= \text{Tr}_{X,Y} (\Pi_{i,j} \left| \psi_f \rangle\langle \psi_f \right| \Pi_{i,j}) .
\end{align}
%
From the reduced density matrix, the following two amplitudes for the measurement results can be extracted
%
\begin{align}
    a_0(x_i,y_j) & = \frac{\left\langle 0 |\rho_{i,j}| 0 \right\rangle}{\text{Tr}(\rho_{i,j})}  \\
    a_1(x_i,y_j) & = \frac{\left\langle 1 |\rho_{i,j}| 1 \right\rangle}{\text{Tr}(\rho_{i,j})} .
\end{align}
%
Currently available quantum platforms are still subject to a substantial level of noise and extracting the exact distance from amplitude is still a difficult task. In order to reduce noise we average the measurement results over different states of the ancillary qubit, thus the measured Hamming distance between the input vector $\left| x_i \right\rangle$ and cluster vector $\left| y_j \right\rangle$ is
%
\begin{align}
    d_{i,j}^H & \propto 1 - \frac{1}{2}(a_0(x_i,y_j) + (1-a_1(x_i,y_j))) .
\end{align}
%
The Hamming distance measured in this way is bounded $0 \leq d_{i,j}^H \leq 1$,
where  $0$ indicates that $x_i$ and $y_j$ are identical and $1$ means they are the completely opposite in terms of their pairwise binary coordinates.

>>>>

\begin{figure}[t]
	\includegraphics[width=\columnwidth]{qcircuit.eps}
  \caption{%
		Quantum circuit for the quantum parallelized Hamming distance calculation between all pairs of binary vectors from two sets ${X}$ and ${Y}$.
		First, we encode information about pairwise different qubits in a quantum state of the $X$-register by applying the CNOT gates.
		Second, the Hamming distance values are extracted to the amplitudes of superposition with the control phase rotation (see Eq.~(\ref{eq:controled_phase_rotation})) and Hadamard gates.
		Finally, a quantum space of the $X$-register is returned to the initial basis for information retrieval.
	}
	\label{fig:qcircuit}
\end{figure}

\begin{figure}[t]
  \includegraphics[width=\columnwidth]{distance_matrix.eps}
  \caption{%
    Result of tomography of all vector pairs that differ by up to one bit --- neighbors.
    Distance values less than 1.7 are marked black.
    Distance calculated by $\bar d_{ij} = (d_{ij} + d_{ji}) / 2$,
    where $d_{ij}$ is extracted from amplitudes by~(\ref{eq:amplitudes2distance}).
    Amplitudes obtained on the IBM quantum backend named ``ibm\_sherbrook'' with only 8192 shots.
    Question marks and crosses indicate missing and erroneous results compared to theoretical calculations.
    The accuracy compared to classical simulation is better than 97 $\%$.
  }
  \label{fig:distance_matrix}
\end{figure}

@@

Measuring the Hamming distance of a particular pair of input vectors $\left| x_i \right\rangle$ and cluster vectors $\left| y_j \right\rangle$ consists of extracting the relevant amplitude from the subspace that those states form,
this can be done using the following projection operator
%
\begin{align}
\Pi_{i,j} = &\left| x_i \rangle\langle x_i \right| \otimes \left| y_j \rangle\langle y_j \right| \otimes I .
\end{align}
%
Using the above projection operator, the subspace of the Hilbert space formed by a particular pair of input and cluster vectors can be traced out as
%
\begin{align}
    \rho_{i,j} &= \text{Tr}_{X,Y} (\Pi_{i,j} \left| \psi_f \rangle\langle \psi_f \right| \Pi_{i,j}) .
\end{align}
%
From the reduced density matrix, the following two amplitudes for the measurement results can be extracted
%
\begin{align}
    a_0(x_i,y_j) & = \frac{\left\langle 0 |\rho_{i,j}| 0 \right\rangle}{\text{Tr}(\rho_{i,j})}  \\
    a_1(x_i,y_j) & = \frac{\left\langle 1 |\rho_{i,j}| 1 \right\rangle}{\text{Tr}(\rho_{i,j})} .
\end{align}
%
Currently available quantum platforms are still subject to a substantial level of noise and extracting the exact distance from amplitude is still a difficult task. In order to reduce noise we average the measurement results over different states of the ancillary qubit, thus the measured Hamming distance between the input vector $\left| x_i \right\rangle$ and cluster vector $\left| y_j \right\rangle$ is
%
\begin{equation}\label{eq:amplitudes2distance}
    d_{i,j}^H \propto 1 - \frac{1}{2}(a_0(x_i,y_j) + (1-a_1(x_i,y_j))).
\end{equation}
%
The Hamming distance measured in this way is bounded $0 \leq d_{i,j}^H \leq 1$,
where  $0$ indicates that $x_i$ and $y_j$ are identical and $1$ means they are the completely opposite in terms of their pairwise binary coordinates.
